---
title: "VTPEH 6270 - Check Point 03"
subtitle: "Data Set Selection & R markdown Report Outline"
author: "Yi(Eve) Zhang"
date: "2026-02-18"
editor_options:
  chunk_output_type: console
output:
  pdf_document:
    toc: true
    toc_depth: 1
    number_sections: true
urlcolor: blue
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

```

# Research Question and Background

## Research Question
In 2023, how does the prevalence of arthritis among adults vary across U.S. states?

## Scientific Plausibility and Context
Arthritis is one of the most common chronic conditions among U.S. adults and a major contributor to pain, disability, and reduced quality of life [@NCHS497_2024].  
Adults with arthritis are generally less physically active than those without, and many report engaging in only low-impact activities such as walking, which highlights the functional impact of the condition [@CDC_MMWR_mm7040a3].  
Counties with higher arthritis prevalence and arthritis-attributable pain outcomes show significant spatial clustering in the United States, indicating that geographic variation is not random [@Sun2024_GeographyArthritisPain].  
Therefore, it is scientifically plausible that adult arthritis prevalence varies across U.S. states in 2023, reflecting differences in the distribution of higher- and lower-burden counties within states.

# Data Preparation

The analytic dataset includes a continuous outcome (county-level crude arthritis prevalence) and a categorical grouping variable (state), enabling grouped summaries and visualization.


## Import and cleaning

```{r, results='hide'}
library(readr)
library(dplyr)
library(stringr)
library(janitor)
library(forcats)
library(ggplot2)

#import data
setwd("/Users/eve/Desktop/V270")
places <- read_csv("PLACES__Local_Data_for_Better_Health,_County_Data,_2025_release_20260127.csv") |>
  clean_names()
#checking column names
names(places)
#checking data_value_type
places |>
  distinct(data_value_type) |>
  arrange(data_value_type) |>
  print(n = 50)

# Subset to analytic dataset for the research question
arthritis_2023 <- places |>
  filter(
    year == 2023,
    measure == "Arthritis among adults",
    data_value_type == "Crude prevalence"
  ) |>
  filter(!is.na(data_value)) |>
  select(
    state_desc, location_name,
    year, measure, data_value_type,
    data_value, low_confidence_limit, high_confidence_limit
  )

# Quick checks
arthritis_2023 |>
  summarise(
    n_rows = n(),
    n_states = n_distinct(state_desc),
    min = min(data_value),
    max = max(data_value)
  )
```

## Derived Variables
```{r}
arthritis_2023 <- arthritis_2023 |>
  mutate(`arthritis_prop` = data_value / 100)

summary(arthritis_2023$arthritis_prop)

```
The research question compares crude arthritis prevalence (`data_value`) across states (`state_desc`) directly. I converted the crude prevalence from percent to proportion to facilitate interpretation (`arthritis_prop = data_value/100`).


## Categorization and Discretization

No discretization was needed because the grouping variable `state_desc` is categorical (U.S. state), which supports grouped summaries and visualization; the outcome `data_value` is continuous (county-level crude prevalence).

# Grouped Summary Statistics
## Stratified Analysis
### Table 1

```{r}

library(knitr)
library(kableExtra)
```


```{r}

table_AB <- arthritis_2023 |>
  filter(!state_desc %in% c("United States")) |>
  group_by(state_desc) |>
  summarise(
    n_counties = n(),
    mean_pct = mean(data_value, na.rm = TRUE),
    sd_pct = sd(data_value, na.rm = TRUE),
    median_pct = median(data_value, na.rm = TRUE),
    iqr_pct = IQR(data_value, na.rm = TRUE),
    min_pct = min(data_value, na.rm = TRUE),
    max_pct = max(data_value, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(desc(median_pct)) |>
  mutate(across(where(is.numeric), ~ round(.x, 1)))

table_AB |>
  kable(
    format = "latex",
    booktabs = TRUE,
    longtable = TRUE,
    caption = "table_AB: Descriptive statistics for county-level crude arthritis prevalence (percent), stratified by state, 2023 (PLACES)",
    col.names = c("State", "N counties", "Mean", "SD", "Median", "IQR", "Min", "Max")
  ) |>
  kable_styling(latex_options = c("repeat_header"), font_size = 9)
```



### Table 2



